Linear interaction of treatment with baseline risk was sufficient for predicting individualized benefit

### 2021-09-23 submission/manuscript.rmd
add model selection rates for adaptive

---
title: |
 Linear interaction of treatment with baseline risk was sufficient for predicting
 individualized benefit
abstract: |
 **Objective:** To compare different risk-based methods predicting individualized
 treatment effects in RCT simulations. **Study Design and Setting:** We simulated
 data using diverse assumptions for a baseline prognostic index of risk (PI) and
 the shape of its intereraction with treatment (none, linear or quadratic). In
 each sample we predicted absolute benefit using: models with the PI and a
 constant relative treatment effect; models including an interaction of treatment
 with the PI; stratification in quarters of the PI; nonlinear transformations of
 the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach
 using Akaike’s Information Criterion. We evaluated predictive performance using
 root mean squared error and measures of discrimination and calibration for
 benefit. Starting from a base case scenario (sample size 4,250, treatment odds
 ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect
 strength, and the PI's discriminative ability. **Results:** Models including a
 PI by treatment interaction performed better under most simulation settings.
 Flexible models required larger sample sizes and higher AUC of the PI to
 outperform the linear interaction model. The adaptive approach tended to select
 simpler models with smaller sample sizes and/or worse prediction performance.
 **Conclusion:** Under most circumstances, a model with a linear interaction of
 the PI with treatment is the optimal risk-based approach to predicting absolute
 treatment benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

@@

The model including a constant relative treatment effect had the lowest median
RMSE in scenarios with a true constant relative treatment effect (OR = 0.8, N =
`r 4250` and AUC = 0.75) or with moderate relative deviations (Figure
\ref{fig:rmsebase}; Panel A). However, when we considered strong linear and
quadratic deviations from a constant relative treatment effect the linear
interaction model performed best (Figure \ref{fig:rmsebase}; Panels B and C).
Only in the case of strong quadratic deviations models including RCS (3 knots)
performed equally well to the linear interaction method. Increasing the number
of knots in RCS resulted in higher RMSE across all scenarios. The adaptive
approach performed very similarly to the best performing model in each scenario.

@@

When we increased the AUC of the true prediction model to 0.85, models including
RCS smoothing had the lowest RMSE in the presence of strong quadratic deviations
from the base case of true constant relative treatment effects (Figure
\ref{fig:rmseauc}; Panel D). However, with milder deviations, the linear
interaction model had the lowest RMSE with the RCS smoothing methods (3 knots)
being a close second (Figure \ref{fig:rmseauc}; Panels B and C). Increasing the
number of knots of RCS smoothing resulted in increased RMSE, which was less
pronounced in the case of strong quadratic deviations.

>>>>

---
title: |
 Linear interaction of treatment with baseline risk was sufficient for predicting
 individualized benefit
abstract: |
 **Objective:** To compare different risk-based methods predicting individualized
 treatment effects in RCT simulations. **Study Design and Setting:** We simulated
 data using diverse assumptions for a baseline prognostic index of risk (PI) and
 the shape of its intereraction with treatment (none, linear or quadratic). In
 each sample we predicted absolute benefit using: models with the PI and a
 constant relative treatment effect; models including an interaction of treatment
 with the PI; stratification in quarters of the PI; nonlinear transformations of
 the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach
 using Akaike’s Information Criterion. We evaluated predictive performance using
 root mean squared error and measures of discrimination and calibration for
 benefit. Starting from a base case scenario (sample size 4,250, treatment odds
 ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect
 strength, and the PI's discriminative ability. **Results:** Models including a
 PI by treatment interaction performed better under most simulation settings.
 Flexible models required larger sample sizes and higher AUC of the PI to
 outperform the linear interaction model. The adaptive approach tended to select
 simpler models with smaller sample sizes and/or worse prediction performance.
 **Conclusion:** Under most circumstances, a model with a linear interaction of
 the PI with treatment is the optimal risk-based approach to predicting absolute
 treatment benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[pagewise, left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

\doublespacing 
\linenumbers

@@

The model including a constant relative treatment effect had the lowest median
RMSE in scenarios with a true constant relative treatment effect (OR = 0.8, N =
`r 4250` and AUC = 0.75) or with moderate relative deviations (Figure
\ref{fig:rmsebase}; Panel A). However, when we considered strong linear and
quadratic deviations from a constant relative treatment effect the linear
interaction model performed best (Figure \ref{fig:rmsebase}; Panels B and C).
Only in the case of strong quadratic deviations models including RCS (3 knots)
performed equally well to the linear interaction method. Increasing the number
of knots in RCS resulted in higher RMSE across all scenarios. The adaptive 
approach selected a linear interaction model 35\% of the time in the case of
true strong linear deviations, while the majority of the time (56\%) it selected
a simpler constant treatment effect model (Supplementary Table S7). Similarly, 
in the case of true strong quadratic deviations, the adaptive approach selected 
the RCS (3 knots) model 29\% of the time, while the simpler linear interaction 
model was selected 68\% of the time.

@@

When we increased the AUC of the true prediction model to 0.85, models including
RCS smoothing had the lowest RMSE in the presence of strong quadratic deviations
from the base case of true constant relative treatment effects (Figure
\ref{fig:rmseauc}; Panel D). However, with milder deviations, the linear
interaction model had the lowest RMSE with the RCS smoothing methods (3 knots)
being a close second (Figure \ref{fig:rmseauc}; Panels B and C). Increasing the
number of knots of RCS smoothing resulted in increased RMSE, which was less
pronounced in the case of strong quadratic deviations. With increasing 
prediction AUC we saw an increased tendency of the adaptive approach to select
more complex models compared to the base case scenario. In the case of strong 
linear deviations, the linear interaction model was selected the majority of the
time (63\%), while the simpler constant treatment effect model was selected 26\%
of the time (Supplementart Table S7). However, the increased RMSE of the
constant treatment effect model inflated the the RMSE of the linear interaction
model. In the case of strong quadratic deviations the RCS (3 knots) model was
selected 84\% of the time.

### 2021-09-23 submission/manuscript.tex
add model selection rates for adaptive

\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\usepackage{amsmath}
\doublespacing
\usepackage[left]{lineno}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{booktabs}
\date{}
\newcommand\given[1][]{\:#1\vert\:}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

The model including a constant relative treatment effect had the lowest
median RMSE in scenarios with a true constant relative treatment effect
(OR = 0.8, N = 4,250 and AUC = 0.75) or with moderate relative
deviations (Figure \ref{fig:rmsebase}; Panel A). However, when we
considered strong linear and quadratic deviations from a constant
relative treatment effect the linear interaction model performed best
(Figure \ref{fig:rmsebase}; Panels B and C). Only in the case of strong
quadratic deviations models including RCS (3 knots) performed equally
well to the linear interaction method. Increasing the number of knots in
RCS resulted in higher RMSE across all scenarios. The adaptive approach
performed very similarly to the best performing model in each scenario.

>>>>

\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\usepackage{amsmath}
\doublespacing
\usepackage[pagewise, left]{lineno}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{caption}
\usepackage{booktabs}
\date{}
\newcommand\given[1][]{\:#1\vert\:}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

The model including a constant relative treatment effect had the lowest
median RMSE in scenarios with a true constant relative treatment effect
(OR = 0.8, N = 4,250 and AUC = 0.75) or with moderate relative
deviations (Figure \ref{fig:rmsebase}; Panel A). However, when we
considered strong linear and quadratic deviations from a constant
relative treatment effect the linear interaction model performed best
(Figure \ref{fig:rmsebase}; Panels B and C). Only in the case of strong
quadratic deviations models including RCS (3 knots) performed equally
well to the linear interaction method. Increasing the number of knots in
RCS resulted in higher RMSE across all scenarios. The adaptive approach
selected a linear interaction model 35\% of the time in the case of true
strong linear deviations, while the majority of the time (56\%) it
selected a simpler constant treatment effect model (Supplementary Table
S7). Similarly, in the case of true strong quadratic deviations, the
adaptive approach selected the RCS (3 knots) model 29\% of the time,
while the simpler linear interaction model was selected 68\% of the
time.
